Wireless Automobile Detection, License Plate Processing, and Data Availability Network Proposal

### 2014-03-15 proposal/proposal.tex
Added Base Station Section

The parking lots at the Colorado School of Mines (CSM) can be a source of frustration for students and faculty members. Because space on campus is limited, the lots are often unable to accommodate everyone who needs to use them. Students need to decide which lot to park in before they go to class; however, knowing which lots have free spaces usually requires driving through the lots looking for a space. The extra time spent searching one or many lots can make students or faculty late for class. If campus members could find out which lots had available spaces before they arrived, much time and frustration could be saved. The goal of this project is to design and deploy a working parking lot monitoring system that will be the first step towards a better parking experience on the CSM campus. The system will keep track of both the number of vehicles in a parking lot and the license plate numbers of those vehicles. The information will be available online to system administrators and CSM campus members. There are two main parts to this project:

@@

The idea behind the first part is that a student can get on their smartphone, tablet, or laptop before heading to class, see a nicely-formatted list/map of the lots on campus, and decide where to park based on how full each lot is. Administrators will monitor the information and make sure it is correct, updating the web application as needed with new statistics, lists and maps as more lots are added to the system. The Association for Computing Machinery group (ACMx) at CSM has already started working on a system for monitoring lot capacities (see description of the SmartLots project in section 1.2). We will collaborate closely with the ACMx group, integrating our work with theirs to create and deploy a working system by the end of this semester (May 2014).

@@

The goal of this project is to work with the ACMx group to deploy the system in two parking lots on campus: the CTLM upper lot and CTLM lower lot, which are the two closest parking lots to the CTLM building on the CSM campus. The system itself will consist of five main subsystems, which are described in detail in section 2.

\subsection{Related Work}
The ACMx group at CSM has designed a system named SmartLots to track the number of vehicles in a parking lot. Their system is a wireless sensor network consisting of sensor nodes and a central basestation. The sensor nodes are based on the Arduino Fio microcontroller platform equipped with triaxial magnetometers and XBee Pro radios. The base station is one small, commercially available Raspberry Pi Linux computer. The information collected by the sensor nodes is transmitted to the base station and forwarded to a server running a small web application. The system is described in \cite{stillwell2013} and \cite{parkingWiki}. \cite{stillwell2013} describes an implementation of a system to detect automobiles entering and exiting a parking lot using the Arduino Fio and the Raspberry Pi. This system uses the IEEE 802.15.4 communications standard to communicate automobile detection data from the sensor nodes to the base station. \cite{parkingWiki} describes the ongoing status of the ACMx project. At the time of writing this proposal, the ACMx team has created a website that will display the status of the CTLM upper and lower lots once the monitoring system has been deployed. \cite{parkingWiki} is updated periodically with new information about project progress. The project described in this proposal will augment and extend the ACMx project while collaborating closely with Stillwell (ACMx president and author of \cite{stillwell2013}), with the goal of a real world deployment by May 2014.

\section{Proposed Work}
\subsection{Overview}
The project we are proposing will be an extension of the ACMx SmartLots project. We will collaborate closely with Stillwell to create and deploy the first working parking lot monitoring system on the CSM campus. Our system will consist of several components, each of which will be either an extension of a current SmartLots system or an entirely new subsystem that will be cohesively interfaced. Figure \ref{fig:system} shows the overall architecture of our system.

@@

The system will rely on three main components:
\begin{itemize}
\item Detection Subsystem
\item Central Base Station
\item Server
\end{itemize}
The detection subsystem will be made up of a small number of sensor nodes, placed strategically at the parking lot entrances. Each sensor node will contain an Arduino Fio equipped with a triaxial magnetometer that will be used to detect when a car passes the sensor. The work to determine the hardware configuration and software for the Arduino Fios has already been done by the ACMx group. Our contribution will be to also equip each node with a Raspberry Pi mini-computer and camera, which will take pictures of the license plates of cars as they pass by. When a car passes a sensor node, the magnetometer will detect the car, and the Arduino Fio will send an interrupt to the Raspberry Pi via the XBee radio, waking it up so it can take a picture of the license plate.

@@

\subsection{Automobile Detection}
The automobile detection subsystem provides a means for detecting ingress and egress of vehicles from a parking lot. This subsystem is to be placed on the side of the road next to each parking lot entrance and exit. Automobiles will be detected using a magnetometer which perceives the induced change in the local magnetic field as the metallic structure of the vehicle passes by, as described in \cite{stillwell2013}. The automobile detection subsystem will be based around the commercially available Arduino Fio 16-bit  platform which utilizes the ubiquitous Atmel ATMEGA328p microcontroller. When an automobile is detected by the Arduino Fio, a packet is sent through the onboard XBee to the central basestation (described in Section 2.4) containing the direction of the passing vehicle and a hardware interrupt is sent to the license plate recognition system to awaken it from sleep.

@@

The use of a magnetometer ensures that solely automobiles and other roadway vehicles which such as motorcycles are detected while pedestrians and particulate deposits, such as snow, are ignored. As in \cite{stillwell2013}, we shall use the commercially available HMC5883 triaxial magnetometer as it provides an adequate balance of sensitivity and low power consumption.

\subsubsection{Tests and Analyses}
Work on the automobile detection subsystem will involve a variety of tests and analyses to ensure system and data integrity. Power consumption analyses will be conducted on the automobile detection and license plate image acquisition subsystem to identify power usage per main component (Figure~\ref{fig:autodetect}) while the mote is in different states, such as sleeping and detecting vehicles. Such an analysis will allow us to allocate how much time the mote is allowed to spend in each state and determine the required lithium polymer battery capacity.

@@

\subsubsection{Interface with Basestation Raspberry Pi}
The recognition and extraction of license plate characters will not directly interface with the Arduino Fio or Acquisition Raspberry Pi that takes the images of automobiles. Instead, it is assumed all communications will come through the Basestation Raspberry Pi, and that the Basestation Raspberry Pi will upload received images to the ACMx Linux server (See section 2.6.1 about the HTTP interface).

@@

\subsubsection{HTTP Interface}
The Basestation Raspberry Pi will need some way to communicate its data with the web application/MySQL database. The simplest way to meet this need is to create an HTTP interface on the server. We will need this interface to handle requests from the front end of the web application anyway, and so it makes sense to also allow the base station to communicate with our server through HTTP requests. We will create a REST-ful HTTP interface that corresponds closely to the database schema and the front-end of the webpage, and the base station will send all its data to the application via the same interface. For example, if an image of a license plate is detected, the base station would send an HTTP POST request to 'acmxlabs.org/parking/licenseImage' (actual urls may change when we start designing the interface), and the server would receive the request and send the image to the ALPR software, as described in section 2.5. When the web application needs to get the data associated with a license plate, it will send an HTTP GET request to 'acmxlabs.org/parking/licensePlate/123ABC'.

\subsubsection{Viewing License Plate Data}
As mentioned in section 2.5, the license plate images will be translated from an image into a string of plain text, and stored in the MySQL database on the ACMx server. One of the main functions of the web interface will be the display of this data to administrators. The page that displays the license plate data will have a simple interface which will allow administrators to filter the data by date/time, lot, and license plate number. For example, an administrator should be able to type in a specific license plate number and see all the dates/times that plate has been seen, and in which lot. Alternatively, an admin will be able to enter a date range and lot(s) and see all the license plates that entered/exited the lot(s) during that date range. For now we are only interested in a proof-of-concept type interface that shows that we have the capability to display the collected plate data in a useful format; in the future it could be incredibly useful to cross reference our data with the campus parking department's database to check if a car has entered a lot without a permit.

\subsubsection{Validating and Adjusting Lot Capacity Data}
We expect that our first deployment of this system may not be 100\% accurate at detecting all vehicles that enter and exit lots. Therefore, as mentioned in section 2.4, we will be collecting periodic overhead images from a camera mounted on the Basestation Raspberry Pi. These images will serve as a visual representation of how many cars are actually in the parking lot at the time the image was taken. Administrators should be able to periodically compare this image to the collected data from the sensor network to make sure the numbers being computed by the network accurately represent the number of cars in the lot. As previously mentioned, the deployment of this project will only be on two campus parking lots, and we may only have one overhead camera to start off with, although ideally we would have two - one for each lot. The web interface will allow administrators to pull up the image for any lot with an overhead camera and see the numbers for that lot right next to the image. If any discrepancies are found, the administrator should be able to overwrite the count in the database to reflect reality. This interface will be extremely useful in helping to calibrate the system. For example, if the sensors are consistently reporting less cars in the lot than the image shows, we can fine-tune the collection process to make sure fewer cars slip through unnoticed. Future work in this area will include an automated version of this process, where a computer program will compare the image to the sensed data and make any necessary corrections.

\subsubsection{User Accounts}
As the site currently stands, there is no way to restrict certain data to administrators. We would like to create a simple user log-in system that allows general users to view the main parts of the site (maps of lots, state of each lot, stats and trends) without having to log in, and only allows access to the administrator side of the site (2.6.1 and 2.6.2, among other parts) if the user has a valid administrator username and password. This part of the project will be to add this simple security system to the site, and to create an administrator interface that allows top administrators to create, update, and delete other administrators.

@@

The central basestation, as described in Section 2.4, will be located on the second level of the CTLM building. Several potential locations exist which allow for a clear line-of-sight to all automobile detection subsystems, one of which is detailed in Figure \ref{fig:deployment}. These locations also provide internet access to the Basestation Raspberry Pi through ethernet ports.

>>>>

The parking lots at the Colorado School of Mines (CSM) can be a source of frustration for students and faculty members. Because space on campus is limited, the lots are often unable to accommodate everyone who needs to use them. Students need to choose a lot before they go to class; however, knowing which lots have free spaces usually requires driving through the lots looking for a space. The extra time spent searching one or many lots can make students or faculty late for class. If campus members could find out which lots had available spaces before they arrived, much time and frustration could be saved. The goal of this project is to design and deploy a working parking lot monitoring system that will be the first step towards a better parking experience on the CSM campus. The system will keep track of both the number of vehicles in a parking lot and the license plate numbers of those vehicles. The information will be available online to system administrators and CSM campus members. There are two main parts to this project:

@@

The idea behind the first part is that a student can get on their smartphone, tablet, or laptop before heading to class, see a nicely-formatted list/map of the lots on campus, and decide where to park based on how full each lot is. Administrators will monitor the information and make sure it is correct, updating the web application as needed with new statistics, lists and maps as more lots are added to the system. The Association for Computing Machinery group (ACMx) at CSM has already started working on a system for monitoring lot capacities (see section 1.2). We will collaborate closely with the ACMx group, integrating our work with theirs to create and deploy a working system by the end of this semester (May 2014).

@@

The goal of this project is to work with the ACMx group to create and deploy the system in two parking lots on campus by the end of the semester. See section 2.7 for information about the deployment. The system itself will consist of several subsystems, which are described in detail in section 2.

\subsection{Related Work}
The ACMx group at CSM has designed a system named SmartLots to track the number of vehicles in a parking lot. Their system is a wireless sensor network consisting of sensor nodes and a central base station. The sensor nodes are based on the Arduino Fio microcontroller platform equipped with triaxial magnetometers and XBee Pro radios. The base station is one small, commercially available Raspberry Pi Linux computer. The information collected by the sensor nodes is transmitted to the base station and forwarded to a server running a small web application. The system is described in \cite{stillwell2013} and \cite{parkingWiki}. \cite{stillwell2013} describes an implementation of a system to detect automobiles entering and exiting a parking lot using the Arduino Fio and the Raspberry Pi. This system uses the IEEE 802.15.4 communications standard to communicate automobile detection data from the sensor nodes to the base station. \cite{parkingWiki} describes the ongoing status of the ACMx project. At the time of writing this proposal, the ACMx team has created a website that will display the status of the CTLM upper and lower lots once the monitoring system has been deployed. \cite{parkingWiki} is updated periodically with new information about project progress. The project described in this proposal will augment and extend the ACMx project while collaborating closely with Stillwell (ACMx president and author of \cite{stillwell2013}), with the goal of a real world deployment by May 2014.

\section{Proposed Work}
\subsection{Overview}
The project we are proposing will be an extension of the ACMx SmartLots project. We will collaborate closely with Stillwell and the ACMx group to create and deploy the first working parking lot monitoring system on the CSM campus. Our system will consist of several components, each of which will be either an extension of a current SmartLots system or an entirely new subsystem that will be integrated into the existing system. Figure \ref{fig:system} shows the overall architecture of our system.

@@

The system will rely on three main components:
\begin{enumerate}
\item Detection Subsystem
\item Central Base Station
\item Server
\end{enumerate}
The detection subsystem will be made up of a small number of sensor nodes, placed strategically at the parking lot entrances. Each sensor node will contain an Arduino Fio equipped with a triaxial magnetometer that will be used to detect when a car passes the sensor. The work to determine the hardware configuration and software for the Arduino Fios has already been done by the ACMx group. Our contribution will be to also equip each node with a Raspberry Pi mini-computer and camera, which will take pictures of the license plates of cars as they pass by. When a car passes a sensor node, the magnetometer will detect the car, and the Arduino Fio will send an interrupt to the Raspberry Pi via the XBee radio, waking it up so it can take a picture of the license plate.

@@

\subsection{Automobile Detection}
The automobile detection subsystem provides a means for detecting ingress and egress of vehicles from a parking lot. This subsystem is to be placed on the side of the road next to each parking lot entrance and exit. Automobiles will be detected using a magnetometer which perceives the induced change in the local magnetic field as the metallic structure of the vehicle passes by, as described in \cite{stillwell2013}. The automobile detection subsystem will be based around the commercially available Arduino Fio 16-bit  platform which utilizes the ubiquitous Atmel ATMEGA328p microcontroller. When an automobile is detected by the Arduino Fio, a packet is sent through the onboard XBee Pro radio to the central base station (described in Section 2.4). The packet will  contain the direction of the passing vehicle. At the time of detection,  a hardware interrupt is also sent to the license plate recognition system to awaken it from sleep.

@@

The use of a magnetometer ensures that automobiles and other roadway vehicles such as motorcycles are detected, while pedestrians and particulate deposits, such as snow, are ignored. As in \cite{stillwell2013}, we shall use the commercially available HMC5883 triaxial magnetometer as it provides an adequate balance of sensitivity and low power consumption.

\subsubsection{Tests and Analyses}
Because the ACMx group has already written the necessary software and designed a working configuration for detecting vehicles with the Arduino Fios, the bulk of our work on the automobile detection subsystem will involve a variety of tests and analyses to ensure system and data integrity. Power consumption analyses will be conducted on the automobile detection and license plate image acquisition subsystem to identify power usage per main component (Figure~\ref{fig:autodetect}) while the mote is in different states, such as sleeping and detecting vehicles. Such an analysis will allow us to allocate how much time the mote is allowed to spend in each state and determine the required lithium polymer battery capacity.

@@

\subsubsection{Interface with Base Station Raspberry Pi}
The recognition and extraction of license plate characters will not directly interface with the Arduino Fio or Acquisition Raspberry Pi that takes the images of automobiles. Instead, it is assumed all communications will come through the base station Raspberry Pi, and that the base station Raspberry Pi will upload received images to the ACMx Linux server (See section 2.6.1 about the HTTP interface).

@@

\subsubsection{HTTP Interface}
The base station Raspberry Pi will need some way to communicate its data with the web application/MySQL database. The simplest way to meet this need is to create an HTTP interface on the server. We will need this interface to handle requests from the front end of the web application anyway, and so it makes sense to also allow the base station to communicate with our server through HTTP requests. We will create a REST-ful HTTP interface that corresponds closely to the database schema and the front-end of the webpage, and the base station will send all its data to the application via the same interface. For example, if an image of a license plate is detected, the base station would send an HTTP POST request to 'acmxlabs.org/parking/licenseImage' (actual urls may change when we start designing the interface), and the server would receive the request and send the image to the ALPR software, as described in section 2.5. When the web application needs to get the data associated with a license plate, it will send an HTTP GET request to 'acmxlabs.org/parking/licensePlate/123ABC'.

\subsubsection{Viewing License Plate Data}
As mentioned in section 2.5, the license plate images will be translated from an image into a string of plain text, and stored in the MySQL database on the ACMx server. One of the main functions of the web interface will be to display this data to administrators. The page that displays the license plate data will have a simple interface which will allow administrators to filter the data by date/time, lot, and license plate number. For example, an administrator should be able to type in a specific license plate number and see all the dates/times that plate has been seen, and in which lot. Alternatively, an administrator will be able to enter a date range and lot(s) and see all the license plates that entered/exited the lot(s) during that date range. For now we are only interested in a proof-of-concept type interface that shows we have the capability to display the collected plate data in a useful format; in the future it could be incredibly useful to cross reference our data with the campus parking department's database to check if a car has entered a lot without a permit.

\subsubsection{Validating and Adjusting Lot Capacity Data}
We expect that our first deployment of this system may not be 100\% accurate at detecting all vehicles that enter and exit lots. Therefore, as mentioned in section 2.4, we will be collecting periodic overhead images from a camera mounted on the base station Raspberry Pi. These images will serve as a visual representation of how many cars are actually in the parking lot at the time the image was taken. Administrators should be able to periodically compare this image to the collected data from the sensor network to make sure the numbers being computed by the network accurately represent the number of cars in the lot. As previously mentioned, the deployment of this project will only be on two campus parking lots, and we may only have one overhead camera to start off with, although ideally we would have two - one for each lot. The web interface will allow administrators to pull up the image for any lot with an overhead camera and see the numbers for that lot right next to the image. If any discrepancies are found, the administrator should be able to overwrite the count in the database to reflect reality. This interface will be extremely useful in helping to calibrate the system. For example, if the sensors are consistently reporting less cars in the lot than the image shows, we can fine-tune the collection process to make sure fewer cars slip through unnoticed. Future work in this area will include an automated version of this process, where a computer program will compare the image to the sensed data and make any necessary corrections.

\subsubsection{User Accounts}
As the site currently stands, there is no way to restrict certain data to administrators. We would like to create a simple user log-in system that allows general users to view the main parts of the site (maps of lots, state of each lot, stats and trends) without having to log in, and only allows access to the administrator side of the site (2.6.1 and 2.6.2, among other parts) if the user has a valid administrator username and password. This part of the project will also include an administrator interface that allows top administrators to create, update, and delete other administrators.

@@

The central base station, as described in Section 2.4, will be located on the second level of the CTLM building. Several potential locations exist which allow for a clear line-of-sight to all automobile detection subsystems, one of which is detailed in Figure \ref{fig:deployment}. These locations also provide internet access to the base station Raspberry Pi through ethernet ports.
